Viscous Erosion in Porous Media

### 2018-01-09 docs/methods/erosion_methods.tex
worked on intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Erosion is a fluid-mechanical process that occurs when a fluid alters
the geometry.  Erosion is responsible for a variety of processes, many
of which are geophysical.  For example, sand dunes are longer on the
windward side because of the wind erosion.  Another applications is
erosion caused by groundwater flow, where a slow moving fluid erodes
channels, or conduits, in the earth which creates a heterogeneous media
that has a large effect on the resulting flow.  In both these examples,
there is a feedback process between the flow and the erodible
bodies---the shape of the eroding body affects the flow, while the flow
changes the shape of the eroding body.  This fluid-mechanical feedback
system has been simulated in idealized cases such as a single eroding
two-dimensional body in a uniform flow~\cite{moo-ris-chi-zha-she2013} or
a single mobile eroding three-dimensional body~\cite{mit-spa2016}. To
further develop insight into the effect of two-dimensional erosion, we
consider erosion of a multiply-connected domain
(Figure~\ref{fig:50bodies}), such as a porous media that is typically
used to model groundwater flow.

Because of the scales involved with groundwater flow, we will restrict
our attention to Stokesian dynamics.  For groundwater flow, Darcy's
equation are often used, but this assumes a homogeneous geometry which
is not always a fair assumption.  Instead, we will use the
incompressible Stokes equations to represent the flow, and the
individual pores will be be resolved, resulting in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited as it can resolve complex geometries and the non-negligible
interactions between bodies are resolved, all the while high-order
accuracy and computational efficiency are achieved.  To simulate the
erosion, each body will erode at a rate that is proportional to the
shear stress.  To further visualize the effect of erosion, we compute
the vorticity of the flow (Figure~\ref{fig:50bodies}) which is equal to
the shear stress in the Stokesian limit.  Finally, to evolve the
geometry, a second-order time stepping method that guarantees that the
mesh does not become distorted will be employed.

\paragraph{Contributions} The present study will develop numerical
methods to simulate viscous erosion of multiply-connected
two-dimensional geometries.  By assuming the flow is viscous, we are
assuming the flow is very slow and in small length scales, as is the
case in groundwater flow.  Moreover, by allowing for complex
multiply-connected domains, the effects due to interacting bodies on the
erosion are captured.

Because of the low Reyonlds number assumption, the flow will be assumed
to be Stokesian.  Then, to create high-fidelity and computationally
efficient numerical methods, two fast and accurate methods will be
employed:
\begin{enumerate}
  \item The fluid equations will be solved with spectral accuracy by
  using an efficient boundary integral equation method. Moreover, the
  rate of erosion will be computed and the rate of erosion will be
  computed with spectral accuracy.
  \item Given a rate of erosion, the shape will be integrated in time
  with second-order accuracy.  All spatial derivatives will be computed
  in Fourier space, resulting in spectral accuracy in space.
\end{enumerate}

@@

To model erosion, we the normal velocity is prescribed to be
proportional to the absolute value of the shear stress.  The constant of
proportionality sets the time scale, and we set its value to be 1
\begin{align}
  \dot{\xx}(t) = \abs{\tau} \nn.
  \label{eqn:bodyEvolution}
\end{align}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Erosion is a fluid-mechanical process that occurs when a fluid alters
the geometry.  Erosion is responsible for a variety of processes, many
of which are geophysical.  For example, sand dunes are longer on the
windward side because of the wind erosion.  Another applications is
erosion caused by groundwater flow, where a slow moving fluid erodes
channels, or conduits, in the earth which creates a heterogeneous media
that has a large effect on the resulting flow. Erosion is also important
in the biophysical community. Plaque erosion in arteries is
key to understanding heart disease~\cite{sha2002,
gro-gij-van-fer-hat-van-yua-wen2007}, and the separation of biofilms can
be explained by erosion~\cite{pic-van-hei2000}. In all these examples,
there is a feedback process between the flow and the erodible
bodies---the shape of the eroding body affects the flow, while the flow
changes the shape by fluidizing the eroding body.  This fluid-mechanical
feedback system has been simulated in idealized cases such as a single
eroding two-dimensional body in a uniform
flow~\cite{moo-ris-chi-zha-she2013} or a single mobile eroding
three-dimensional body~\cite{mit-spa2016}. To further develop insight
into the effect of two-dimensional erosion, we consider erosion of a
multiply-connected domain (Figure~\ref{fig:50bodies}), such as a porous
media that is typically used to model groundwater flow.

Because of the scales involved with groundwater flow, we will restrict
our attention to Stokesian dynamics.  For groundwater flow, Darcy's
equation are often used, but this assumes a homogeneous geometry which
is not always a fair assumption.  Instead, we will use the
incompressible Stokes equations to represent the flow, and the
individual pores will be be resolved, resulting in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited as it can resolve complex geometries and the non-negligible
interactions between bodies are resolved, all the while high-order
accuracy and computational efficiency are achieved. To link the body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  To further visualize the effect
of erosion, we compute the vorticity of the flow
(Figure~\ref{fig:50bodies}) which is equal to the shear stress in the
Stokesian limit.  Finally, to evolve the geometry, a second-order time
stepping method that guarantees that the mesh does not become distorted
will be employed.

\paragraph{Contributions} The present study will develop numerical
methods to simulate viscous erosion of multiply-connected
two-dimensional geometries.  By assuming the flow is viscous, we are
assuming the flow is very slow and in small length scales, as is the
case in groundwater flow.  Moreover, by allowing for complex
multiply-connected domains, the effects due to interacting bodies on the
erosion are captured.  We start by authenticating our method on a single
circular body placed symmetrically in a pipe flow.  By artificially
fixing the area of the body as a function of time, we are able to
characterize the long-time behavior of the shape and the resulting flow.
Then, we compute the vorticity of erosion in multiply-connected
geometries to better characterize erosion in porous media.

Because of the low Reynolds number assumption, the flow will be assumed
to be Stokesian.  Then, to create high-fidelity and computationally
efficient numerical methods, two fast and accurate methods will be
employed:
\begin{enumerate}
  \item The fluid equations will be solved with spectral accuracy by
  using an efficient boundary integral equation method. Moreover, the
  rate of erosion will be computed and the rate of erosion will be
  computed with spectral accuracy.
  \item Given a rate of erosion, the shape will be integrated in time
  with second-order accuracy.  All spatial derivatives will be computed
  in Fourier space, resulting in spectral accuracy in space.
\end{enumerate}
We use a quasi-steady assumption so that these two steps can be iterated
to simulate the dynamics.

The effect of erosion on a geometry is quite rich.  For instance, it has
been hypothesized and numerically observed that the erosion leads to a
shear stress that is uniform on the body which results in a uniform
erosion rate.  The shape that has a uniform shear stress is analyzed in
three dimensions by Montenegro-Johnson and Lauga~\cite{mon-lau2015}
which is an extension of the analysis done by Pironneau~\cite{pir1973}
to find a Stokesian drag minimizing shape.  However, the effect of
erosion on a porous media is less understood.  With the methods
developed in this paper, quantities such as the total drag across a
porous region as a function of time can be computed in future work.

A challenge with erosion is that it creates corners at points on the
body where the shear stress vanishes.  While recent work does allow for
integral equation methods to resolve a corner, we employ a different
method.  First, we add a regularization term to the shear stress to
penalize regions of high curvature.  Second, we apply a Gaussian filter
to smooth the absolute value of the shear stress before the body shape
is updated.

@@

To model erosion, we the normal velocity is prescribed to be
proportional to the absolute value of the shear stress.  This choice is
based on experimental work of Ristroph et
al.~\cite{ris-moo-chi-she-zha2012}, but other models for erosion are
possible.  For example, some models use an erosion rate that is
proportional to a power of the magnitude of the shear
stress~\cite{par-izu2000}, while others introduced a shear stress below
which erosion does not occur~\cite{lag2000}.

The constant of proportionality sets the time scale, and we set its
value to be 1
\begin{align}
  \dot{\xx}(t) = \abs{\tau} \nn.
  \label{eqn:bodyEvolution}
\end{align}
